\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-06-13 paper/paper.tex
added natbib for citations

\usepackage{color} % used to mark parts that need to be edited

\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

@@

\begin{document}

@@

% 2.Abschnitt: Bedeutung der Datengrundlage und Einführung von Steuerdaten
Given the importance of the subject a constant reflection about reliability of empirical data seems appropriate. Atkinson (2013:8) observes advances in technology and methodology which improves the core sources of inequality research, the household surveys.  On the other hand the labor intensive and expensive surveys around the world are subject to budget cuts and the instrument itself faces problems in form of low response rates, which affects the assessment of inequality undisputedly. These concerns have led to the search of alternative data sources, which can supplement the established survey data studies. Already Kuznet (1955) used tax data to examine the relationship between economic growth and personal distribution of income. Then it took several decades until Piketty (2001,2003) made the use of tax data fashionable again. Following his approach studies on several countries were conducted (Atkinson and Piketty 2007, 2010). Today, all existing top income tax statistics based time series are collected and accessible through the world top incomes database (Alvaredo et al. 2014). \\

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concering inequality in income. What is known about Switzerland so far? Looking for offical data, three main sources has to be mentioned, which can be considered as de facto offical data sources: EU-SILC, HBS and LIS-data. Figure XY shows the results steming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common ``framework'' to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover timepoints before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm last accessed 21.Mai 2014.} The second important source concering the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on houshold budegts. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistend time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast the the aformentioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data Gornick and Jäntti (2013) found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by Grabka and Kuhn (2012) analyzing the Swiss Household Panel (2000-2009).\footnote{A further offical database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). A further important database on inequality is the GINI database which has been derived from the GINI Country Reports (Nolan et al.,2014). But this dataset doesn't cover Switzerland.}  \\

@@

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study, EU-SILC). However, it is unclear how the bias through non-response affects the overall measure of inequality. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) and a time series for inequality of wealth is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific technics to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we compare our results to results from a relevant Inequality-Survey in Switzerland (HBS) to assess the bias through datasource.

@@

Studies on inequality have to adress several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All this questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss  the advantages and shortcomings of tax data compared to other data sources - namely survey data. Along with the despriction we explain the methods and thechnics we use to construct time-series of inequality-measure for income for Switzerland, which we will present in the result sections. By doing this, we demonstrate the suitability of the FTA Tax Data to assess the development of inequality in Switzerland

\subsection{Tax statistics in Swizterland}

@@

Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the oecd (2013) condense these ideas into the ICW framework (income, consumption and wealth), which is ment to be an internationally agreed framework on micro-level statistics.\footnote{Harmonisation with other international standards was an important objective that guided the work of the Expert Group in developing the ICW Framewrok presented in this publication. Considered main stadards were the System of National Accounts (SNA 2008), the canberra Group Handbook on Household Income Statistics (United Nations 2011), the final report of the 17th International Conference of Labour Statisticians (ILO 2004) and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing (UNECE & Eurostat 2006)} According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recomended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or welth holdings, or vice-versa. But it is also stated (OED 2013:18):"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by colecting and provding a database following ths broader concept of economic well-being. http://www.lisdatacenter.org/our-data/lws-database/} \\

This last statment holds for Swizterland too, although the HBS study is strongly influced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awarness of an assessment of income, consumption and wealth simuntanously is rising, we  focus our analysis on income, which is undoubtly a crucial indicator of economic wellbeeing. However, the Federal Tax Office publishes statistics on income and wealth . But it is not possible to analyse the jointly distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can be understood as mandatory consumptions somehow.

@@

The assessment of income inequality is influenced by the definition of the income itself. Marketincome or disposable income for example differ by substantial meaning and by the expected degree of inequality. Therefore the awarness of the analyzed concept is crucial. Terminology can sligthly differ, while common concepts can be identified (for detailed discussion see: OECD (2013:44), Canberra group (2011:24)). Figure xy shows a stylized framework, which includes a distinction of common income sources \footnote{Income from production of household services for own consumption is excluded because this income is hard to measure and not covered in the FTA tax data} and shows the central steps of redistribution. Within this framework common income definitions are situated. These definitions are contrasted with income concepts, which can be derived out of the FTA Tax data.

\includegraphics[width=\maxwidth]{figure/income_definition}

@@

The central income according to tax data is the taxable income. It includes all reported incomes (income from employment, income from property and received transfers\footnote{Mean-tested benefits are not taxed and therefore not included in Tax data. Income for low income groups are therefore underestimated.}) minus several deductions. It is therefore neither a pre-transfer income  nor a post-transfer income measure. It's rather something inbetween. As the FTA tax statistitcs include some but not all deductions\footnote{The difference between the real Total income and the taxable income are deductions. This includes: professional expenses, travel expenses, interest on debt, Alimonies, Training costs, two-earner deduction,Party Contributions, Payment into pillar 3a, purchases in the pension fund and sideline deductions} it is possible to calculate a sort of "total income". As some deducations can be interpreted as compulsary expenses similar to taxes the step towards Total income is a step away from the income, which can be used for consumption. Similiar when calculating the disposable income out of the taxable income through accouting the reported federal taxes, this is a step towards the income, which is left in the basket for consumption (disposable income). Again it is not a "pure" disposable income, because cantonal, municipal taxes and taxes from churches are missing.   \\

@@

If the goal is to catch the income, which shapes the possibility to consume, equivalence scale are implemented. Equivalence scale followe the idea, that persons living in the same household have the possibility to share costs. To compare the individual economic well-beeing equivalence scales adjust for different household sizes (see OECD 2013,173, Buhmann et al., 1998). Because it is neither possible to calculate the total household income nor the respective householdsize out of the FTA tax data, we put this in this paper outside the brackets. Although knowing, that with doing so we neglect a potential aspect, which is capable of shaping the development of inequality.

% letzter Abschnitt muss noch auf die Brülhart-Daten hinweisen. Damit können wir eine äquivalenzeinkommen betrachten.
% Auch müsste in diesem Abschnitt auf die Probleme der Haushaltsunterscheidung eingegangen werden (oder ist im Abschnitt Population coverage? Dann müsste obiger Punkt auch da erwähnt werden.)

% Streichen
%=======
% 
%>>>>>>> d21c323... completed subsection "concepts on measuring economic resources" and "defining income"

\emph{Population Coverage} \\
All Residents? Working Residents? Single Person vs household/family. Households without income \\
see for details: OCED Framework for Statistics on the Distribution of Houshold income, consumption and wealth (2013) \\

The income of the household generally plays the major role in the living standards attained by its members, so the way earners and non-earners grouped together in households, has to be incorporated into the analysis. Especially when the the focus of interesst lays in unrevealing the role of demographic factors.

\emph{Measuring inequality and concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. Vgl. Allison (1978), Engelhardt (2000), Cowell (2000) oder Hao/Naiman (2010)
%[Zusammenfassung bestehender Überlegung und Hinleitung zu den Masszahlen, die wir verwenden werden (vermutlich Gini und relative distribution methods) \\

neuere Indikatoren; Polaritätsindex: relative distribution methods in the Social Sciences (Handcock/Morris 1999) \\

Gini is mainly used for international comparison. However, other measures are often reported along (gini is more sensitive to changes in the middle of the distribution than to changes in the tails (OECD, 2008:37). Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow better analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading) (Alderson and Doran 2013). \\

Alvaredo (2010) shows formally, how top incomes shares are related to the gini-coefficient. Following his argument, it is crucial to consider top incomes. Leigh (2007) also analyzed the relationship between the Gini (and further inequality measures) and top income shares in a panel of 13 countries (Switzerland is part of the studie). 
%[He also discusses tax data studies in general-> have a closer look] \\
%[We definitely have to argue, why it isnt sufficient to report top income shares, because if it is, there is no need to have another study about Switzerland (national level>Martinez, cantonal level> gorgas and Schaltenegger)]

@@

%[ich stelle mir eine Tabelle mit verschiedenen Dimensionen vor, anhand derer eine Einordnung unterschiedlicher Datenquellen geschieht.]
% Mögliche Vergleichsdimensionen
%- Suitability of cross country comparison
%- Coverage of the population of interesst
%- Implementation of common concepts of economic ressources
%- contingency for intertemporal comparision

%Burkhauser et al. (2009) compare inequality statistics form survey data and tax records to consolidate the findings about recent trends in the USA

@@

%Copied from:
%Alvaredo, Facundo, Anthony B. Atkinson, Thomas Piketty and Emmanuel Saez, The World Top Incomes Database, http://topincomes.g-mond.parisschoolofeconomics.eu/ , 18/12/2013

>>>>

\usepackage{color} % used to mark parts that need to be edited
\usepackage{natbib} %to use citealt citet citep
\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

% 2.Abschnitt: Bedeutung der Datengrundlage und Einführung von Steuerdaten
Given the importance of the subject a constant reflection about reliability of empirical data seems appropriate. \citet[8]{Atkinson 2013:8 FEHLT} observes advances in technology and methodology which improves the core sources of inequality research, the household surveys.  On the other hand the labor intensive and expensive surveys around the world are subject to budget cuts and the instrument itself faces problems in form of low response rates, which affects the assessment of inequality undisputedly. These concerns have led to the search of alternative data sources, which can supplement the established survey data studies. Already \citet{kuznets_economic_1955} used tax data to examine the relationship between economic growth and personal distribution of income. Then it took several decades until \citet{piketty_les_2001, piketty_income_2003,piketty_income_2003-1} made the use of tax data fashionable again. Following his approach studies on several countries were conducted \citep{atkinson_top_2007,atkinson_top_2010}. Today, all existing top income tax statistics based time series are collected and accessible through the world top incomes database \citep{Alvaredo et al. 2014 FEHLT}. \\

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concerning inequality in income. What is known about Switzerland so far? Looking for official data, three main sources has to be mentioned, which can be considered as de facto official data sources: EU-SILC, HBS and LIS-data. Figure XY shows the results stemming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common ``framework'' to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover time points before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm last accessed 21.Mai 2014.} The second important source concerning the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on household budgets. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistent time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast the the aforementioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data \citet{gornick_income_2013} found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by \citet{grabka_evolution_2012} analyzing the Swiss Household Panel (2000-2009).\footnote{A further official database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). For the sake of completeness the database construtcted by \citet{Deininger and Squire 2006 FEHLT} and the World Income Inequality Database (WIID) have to be mentioned. Both datasets do not contain figures  for recent years for Switzerland. A further important database on inequality is the GINI database which has been derived from the GINI Country Reports \citep{nolan_changing_2014}. But this dataset doesn't cover Switzerland.}  \\

@@

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study, EU-SILC). However, it is unclear how the bias through non-response affects the overall measure of inequality. For the US \citet{atkinson_top_2009} estimate that CPS survey data fail to capture about half of the overall increase in inequality measured by the Gini coefficient, a result confirmed by \citet{Alvaredo (2010) FEHLT}. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific techniques to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we compare our results to results from a relevant Inequality-Survey in Switzerland (HBS) to assess the bias through data source.

@@

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All these questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss the advantages and shortcomings of tax data compared to other data sources - namely survey data. We describe and explain methods and techniques needed to construct time-series of income inequality measures for Switzerland. We will present these time-series in the result section and evaluate the use of the FTA Tax Data to assess the development of inequality in Switzerland

\subsection{Tax statistics in Switzerland}

Our data comes from the Swiss Federal Tax Administration (FTA). Federal taxes are collected and documented by the FTA since 1915. Being called a war-tax in the beginning, the federal tax was renamed to crisis levy in 1934, defense-tax in 1939 and is finally known as direct federal tax since 1983. The time frame we were able to collect ranges from 1945 to 2010 including 44 tax periodes. \footnote{Between 1993 and 2003 there is no exact data available for Switzerland because of a system change from taxation assessed in arrears (Praenumerando-System) to taxation assessed on current year income (Postnumerando-System) which was implemented by cantons in different years.}  While the FTA provides data in electronic form since 1973 we collected earlier data by scanning hard copies. Data is available for Switzerland plus all cantons and basically covers every tax unit (individual or household) in Switzerland liable to pay federal taxes. This exempts all tax units with taxable income below a certain threshold (e.g. 15.000 CHF in 2010). Furthermore the FTA differentiates between two groups of tax units, so called normal cases and special cases. A normal case is a tax unit residing in a swiss canton without foreign source income and being liable to taxation all year long. All other tax units and very few that are taxed based on the style of living because they don't work (Pauschalbesteuerte) are special cases. \\

The FTA provides two income measures: taxable income and net income. Net income here is an administrative term and means taxable income plus social deductions (children and supported persons) but not including other deductions like donations or health-care costs. Both measures are designed for taxation purposes which might limit the suitability to measure inequality as we will discuss later.

% Bei den Nullern erw?hnen, ab wann Angaben zur Zahl der Nuller vorliegen und ab wann nicht
% Im Internet hat es Publikationen ab 1947/48
% Es gibt auch Tabellen auf Gemeindeebene
% Die Statitstischen Kennzahlen von Br?lhart erw?hnen?
% Erw?hnen in welcher Form die Informationen publiziert sind. Tabulation by size of income. Not on individual or household level

% http://www.estv.admin.ch/dokumentation/00075/00076/00701/index.html?lang=de
% Kantone und Gemeinden: 2010 bis 1947/1948
% Klassierte Tabellen für Abzüge, Einkommen (Steuerbares Einkommen und Reineinkommen) und Steuererträge
% Normalfälle und Sonderfälle

% Statistsiche Kennzahlen
% Mit Belastung durch die direkte Bundessteuer: 1973/1974 bis 2010
% Mit und ohne Belastung durch die direkte Bundessteuer: 1995/1996 bis 2010
% steuerbares Einkommen/Reineinkommen
% Jeweils Normal und Sonderfälle
% Individual-Daten (nicht klassierte daten)
% äquivalenzeinkommen

@@

Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the \citet{oecd_oecd_2013 UNSICHER} condense these ideas into the ICW framework (income, consumption and wealth), which is ment to be an internationally agreed framework on micro-level statistics.\footnote{Harmonization with other international standards was an important objective that guided the work of the Expert Group in developing the ICW Framework presented in this publication. Considered main standards were the System of National Accounts \citep{sna_system_2008}, the Canberra Group Handbook on Household Income Statistics \citep{united_nations_canberra_2011}, the final report of the 17th International Conference of Labour Statisticians \citep{international_labour_organisation_ilo_final_???? UNSICHER}} and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing \citep{unece_conference_2006}} According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recommended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or wealth holdings, or vice-versa. But it is also stated \citep[18]{oecd_oecd_2013 UNSICHER}:"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by collecting and providing a database following this broader concept of economic well-being. http://www.lisdatacenter.org/our-data/lws-database/} \\

This last statement holds for Switzerland too, although the HBS study is strongly influenced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awareness of an assessment of income, consumption and wealth simultaneously is rising, we  focus our analysis on income, which is undoubtedly a crucial indicator of economic wellbeing. It should be noted, that the Federal Tax Office publishes statistics on income and wealth. But it is not possible to analyze the jointly distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can be understood as mandatory consumptions somehow.

@@

The assessment of income inequality is influenced by the definition of the income itself. Market income or disposable income for example differ by substantial meaning and by the expected degree of inequality. Therefore the awareness of the analyzed concept is crucial. Terminology can slightly differ, while common concepts can be identified (for detailed discussion see: \citet[44]{oecd_oecd_2013 UNSICHER}, \citet[24]{united_nations_canberra_2011}). Figure~\ref{fig:incdef} shows a stylized framework, which includes a distinction of common income sources \footnote{Income from production of household services for own consumption is excluded because this income is hard to measure and not covered in the FTA tax data} and shows the central steps of redistribution, which eventually leads to the disposable income. The income measure, which finaly shapes the possibility to consume. Within this framework common other income definitions are situated. These definitions are contrasted with income concepts, which can be derived out of the FTA Tax data.

\begin{figure}
\centering
\includegraphics[width=\maxwidth]{figure/income_definition}
\caption{Income definitions}
\label{fig:incdef}
\end{figure}

@@

% ACHTUNG: Wir m?ssen innerhalb des Papers einheitlich bleiben. In der Grafik steht jetzt "Total income" oben steht "Net income"

% Man k?nnte es auch konzeptionell auf kantonale Steuerdaten erweitern. Besser w?re es, dies einfach im Ausblick zu erw?hnen

The central income reported through tax statistics is the taxable income. It includes all reported incomes (income from employment, income from property and received transfers\footnote{Realized capital gains are not taxed and Mean-tested benefits are not taxed and therefore not included in Tax data. Income for low income groups are therefore underestimated. However, Piketty and Saez (2001) note that non-taxable social security benefits grew as a share of personal income in the US but find that these changes had nonly a trivial impact on top income shares.}) minus several deductions. It is therefore neither a pre-transfer income  nor a post-transfer income measure. It's rather something in between. As the FTA tax statistics include some but not all deductions\footnote{The difference between the real Total income and the taxable income are deductions. This includes: professional expenses, travel expenses, interest on debt, Alimonies, Training costs, two-earner deduction, Party Contributions, Payment into pillar 3a, purchases in the pension fund and sideline deductions} it is possible to calculate a sort of "total income". As some deductions can be interpreted as compulsory expenses similar to taxes the step towards Total income is a step away from the income, which can be used for consumption. Similar when calculating the disposable income out of the taxable income through accounting the reported federal taxes, this is a step towards the income, which is left in the basket for consumption (disposable income). Again it is not a "pure" disposable income, because cantonal, municipal taxes and taxes from churches, which represent the bulk of taxes in Switzerland, are missing.   \\

@@

\emph{statistical units} \\

The agreed standard on the statistical unit, which should be the base of inequality analysis, are households not individuals \citep[60]{oecd_oecd_2013 UNSICHER}. Indeed it are the individuals, who receive income, own wealth and experience economic wellbeing, but their possibility to do so, is strongly tied to the concept of household. This comprises all persons under the same housing arrangement. The basic underlying assumption for collection data on households level instead of individual level is, that people in the same household share resources and therefore pool their incomes (when two or more earners live together) and/or use the household income to provide the essentials of living for every household member (also non-earing members, like children). Additionally, there are economies of scale when people share living space and commodities and they therefore benefit from the sharing. The compare the individual economic well-being among individuals living in different households usually equivalence scales are used (see \citealt[173]{oecd_oecd_2013 UNSICHER}, \citealt{Buhmann et al., 1998 FEHLT}). \\

In tax data, however, the units are represented according to administrative rules. Tax units therefore neither represent individuals in every case nor true households. Tax units rather represent individuals and couples, but couples, who are married or officially registered. This doesn't imply, that those couples live together, as it is needed to satisfy the definition of a household. On the other hand, is it quit likely that more than one tax units live in the same household (unmarried/unregistered couples, see \citet[99]{muller_vermogenslage_2014}). It is therefore not directly possible to elicit households and household income out of tax data. This might influence the assessment inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing.

% Allenfalls die Br?lhart-Daten erw?hnen und die m?glichkeit der Haushaltsbildung ?ber Abz?ge. Falls wir auch eine Grafik zur Rolle der Haushaltszusammensetzung machen (ich denke schon)
% gibt es studien? Zur Bedeutung von ?quivalenzskalen und zur Bedeutung der Differenzierung von Haushalten und Individuuen?

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. The Gini coefficient is the most known measure and mainly used for international comparison. As it is derived from the lorenz-curve, the quantified amount of inequality can unpretentiously be described in a formal and visual way. Therefore the Gini coefficient is easely interpretable. Furthermore it has several desired statistical properties \citet{engelhardt_modelle_2000}. (1) principle of population: the assessment of inequality is independent of the population size (2) Bresciani-Turroni: the measure is sensitive for changes of income shares, but not for absolute changes (e.g. doubling of all income) (3) weak principle of transfers (pigou-dalton): transfers from richer households lead to a reduction of inequality. However, several drawbacks are reported in the literatur. The most important point is, that the underlying distributional form of the measured inequality is unknown and it is therefore not possibel to see if the measure is driven be a few rich or many poor individuals. This can also be problematic for comparision between countries or over time. In extreme cases two totally different distrubtions share the same gini-coefficient.\\

The recent wave of tax-data studies do not report Gini-coefficents. Rather top income shares are informed on, which are calculated not only with taxdata, but togehter with external sources to produce the population and personal income control totals. This procedure ensures, that the inequality measure is not biased because of non-fillers, who doesn't appear in tax statistics.  Leight (2002:594f) compares top income shares with other inequality measures and ask, if they are a useful measure of inequality in a society. He tries to answer this question empiricaly by comparing measures of inequality based on top income shares with measures of household or familiy inequality. He finds a strong positive relationship, but concludes (P.600): top income shares are far from perfect as a measure of distribution of income across society. Top income shares hence inform not completely on how inequality evolves elsewhere in the distribution. Furthermore, top income shares only weakly satisfy the Pigou-Dalton transfer principle (in contrast to the Gini-Coefficent see above). A transfer from rich to poor will indeed never increase the top income shares, but if the transfer is between invidviduals, who are either both within the top group or both outsie the top group, then the share measure will remain unchanged. \\

Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow better analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading). The polarityindex is developed by \citep{Handock and Morris 1999} allows for this. Recenently this index was applied in the work of \citep{Alderson, Beckfield and Nielsen 2005 FEHLT} and \citep{Alderson and Doran 2013 FEHLT}. This approach is rooted in relative distribution methods. It includes a precise comparison of the shape of two distributions (groups, over time). The main advantage lies herin, that this approche allows to characterize the change, in the sense that it is possible to say, if a polarization occured (reduction of middle class), which equals an increase of inequality and if this change is driven by a change in the upper or lower part of the distribution.  \\

The literature mentions several other metrics with desired properties, but we will not discuss them further (see for example \citet{cowell_chapter_2000} and \citet{hao_assessing_2010}).

% Allenfalls auch auf Quintilvergleiche eingehen und Atkinsonmass

\emph{Population Coverage}
% beim letzten Punkt geht es um Steuerhinterziehung. Fällt das auch noch unter Population coverage?

Often inequality is assessed on national level, which implies, that studies try to cover the whole population of the country of interest. This is a special thorny task for surveys working with samples, because nonresponse is a major source of bias \citep{bethlehem_handbook_2011}.  \citep{korinek_survey_2006} show, that the position in the income distribution influence the probability to participate in a survey. Low income and high income households are more likely to refuse survey response, which leads to an overrepresentation of middle income households. This process can be referred to as the "middleclass bias" \citep{diekmann_empirische_2009}. Missing data in household surveys is therefore not missing at random, which has an impact on the measures of inequality. The magnitude of this bias in Switzerland, however, is unknown. Strategies to handle this kind of bias are discussed in the literature \citep{S?rndal et al. 1992 FEHLT}, but require a register for every unit, that is proportional to income. Currently there exists no such register for Switzerland \citep{ecoplan 2014:43 FEHLT}. Currently used micro datasets, which are used for official publications concerning inequality in Switzerland (SILC and HABE) are furthermore confronted with a constructed coverage problem, because these surveys rely on the phone register, which excludes the households not having a registered connection. \\
%(Problems of collective households, ecoplan 2014:45). Erw?hnen?

The issue of incomplete coverage is less dramatical with tax data. Essentially every permanent resident in Switzerland, who has completed the age of 18 years (respectively 20 years prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). Essentially this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. Most important this includes high net wealth individuals taxed according to their expenditures. Special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units are reported, but not for the years before. Dell et al. (2007) try to estimate the fraction of non-taxed by comparing the reported numbers of tax units to census reports about the number of adult population. Following his etimations this fraction drops from 94\% in 1993/1994 to 63\% back in 1945/46. \\

An other critical issue is the problem of tax evasion, which definitely can bias the assessment of inequality. Alvaredo and Saez (2006) for example regard estimates of Spanish top incomes prior to 1981 as unreliable due to widespread tax evasion. Evasion can occured, when individuals try not to fill tax returns or by misreporting of incomes. In Switzerland non-fillers show up in the tax-statistics either way, as long as they are registered. This person get an imputed income form an older tax return and infomration given by employers. Only non registered non-fillers are not in the records. Therefore non-fillers are a minor problem. Not neglectable is the circumstance, that individuals misreport incomes. \citet{feld_tax_2006} examine the role of tax evasion in Switzerland by calculating the difference of the national accounts measures of primary income and the income reported to the tax authorities. Building on a time series of Pommerehne and Weck-Hannemann (1996) Feld and Frey (2002) and Frey and Feld (2002) extend the series. They can show, that the average level of income tax evasion from 1965 to 1995 varies between 13\% and 35\%. They suggest, that evasion is heavily driven by capital income tax evasion. \\

\emph{Intertemporal comparison}

One advantage of tax data over survey data is their observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HABE data). To be fair, survey data might cover the late periods (2011 and 2012) which are not covered by tax data. The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are consistent for the complete observation period. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.

@@

%[ich stelle mir eine Tabelle mit verschiedenen Dimensionen vor, anhand derer eine Einordnung unterschiedlicher Datenquellen geschieht.]
% Mögliche Vergleichsdimensionen
%- Suitability of cross country comparison
%- Coverage of the population of interesst
%- Implementation of common concepts of economic ressources
%- intertemporal comparision
%- flexibel applications of appropriate inequality measures

\begin{table}[h]
\begin{tabular}{lcccc}
                                   & FTA       & HABE      & EU-SILC   & LIS       \\
Range                              & 1945-2010 & 2000-2011 & 2007-2012 & 1982-2004 \\
Cross country comparison possible? & no        & no        & yes       & yes       \\
Top incomes covered?               & yes       & no        & no        & no        \\
Low incomes covered?               & no        & yes       & yes       & yes       \\
Concepts of economic ressources?   & no        & yes       & ?         & ?         \\
Equivalisation?                    & no        & yes       & ?         &          
\end{tabular}
\end{table}

% Braucht es alle drei Survey-Befragungen
% ich würde nicht dichotomisieren, sondern ordinal bewerten (tief, mittel, hoch)

% Die Punkte kurz beschreiben.
% Zusammenfassung

% Gedankenstütze
%\textbf{Problems with household income surveys}
%\begin{itemize}
%\item Sample data (bias)
%\item comparability between countries and over time (depends on income definition)
%\item short time series
%\end{itemize}

%Burkhauser et al. (2009) compare inequality statistics form survey data and tax records to consolidate the findings about recent trends in the USA
%Studies on problems with survey data See for other countries: \citealt{Siminski et al 2003 FEHLT} (Australia), \citealt{Brewer et al 2008 FEHLT}, UK, \citealt{Burkhauser et al. 2009 FEHLT}, US)

@@

\subsection{Outline of applied methods}

% Diese Abschnitt muss ersetzt werden mit der von uns gewählten Vorgehensweise
